Simple repression draft

### 2020-10-14 doc/resubmission/section_02_means.tex
change noneq for kinetic
implement Rob's changes to main text

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no
transcription for many of these states. This discretization of a potentially
continuous number of promoter states (due to effects such as supercoiling of
DNA~\cite{Chong2014, Sevier2016}  or DNA looping~\cite{Boedicker2013a}) is
analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
continuous molecule conformations into a finite number of
states~\cite{Martins2011}. The models are probabilistic with each state assigned
some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key aspects: what states
to consider and how to compute the probabilities of those states.

@@

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that for all cartoons the
fold-change can be written as a Fermi function 
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~\ref{fig1:means_cartoons}(D), left panel). This implies that all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

@@

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This is \mrm{This is based in one of the
comments from reviewer \# 2. It is empirically true for all models we explored,
but I don't know how to prove that it is a general statement for any
unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way two think about these to terms is as follows: in all kinetic
models shown in Figure~\ref{fig1:means_cartoons}(C) the repressor-bound state
can only be reached from a single repressor-free state. The ratio of these two
states --repressor-bound and adjacent repressor-free state-- must remain the
same for all models, regardless of the details included in other promoter states
if $\Delta F_R$ represents an effective free energy of the repressor binding the
DNA operator. The presence of other states then draws probability density from
the promoter being in either of these two states, making the ratio between the
repressor-bound state and \textit{all} repressor-free states different. The log
difference in this ratio is given by $\log(\rho)$. Since model 1 and model 5 of
Figure~\ref{fig1:means_cartoons}(C) consist of a single repressor-free state,
$\rho$ is then necessarily 1 (See Appendix~\ref{sec:non_bursty} for further
details).

@@

% We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
% is essentially a one-parameter model, the one parameter being $\Delta F_R +
% \log\rho$. By now the reader may be alarmed as to how can we even determine
% $\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
% on the internal structure of $\rho$ itself. A hint is provided by the weak
% promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
% Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
% model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
% approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
% meaning therefore $\rho\approx1$.  This approximation can be well justified on
% the basis of the number of RNAP and $\sigma$ factors per cell
% and the strength of binding of RNAP to
% DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
% not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
% factor? A resolution is offered by an independent inference of $\rho$ in the
% absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
% nonequilibrium model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
% (looking ahead, this is equivalent to model 4
% in Figure~\ref{fig2:constit_cartoons}(A)),
% to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
% determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
% more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

>>>>

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no
transcription for many of these states. This discretization of a potentially
continuous number of promoter states (due to effects such as supercoiling of
DNA~\cite{Chong2014, Sevier2016} or DNA looping \cite{Boedicker2013a}) is
analogous to how the Monod-Wyman-Changeux model of allostery coarse-grains
continuous molecule conformations into a finite number of
states~\cite{Martins2011}. The models are probabilistic with each state assigned
some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key aspects: what states
to consider and how to compute the probabilities of those states.

@@

Figure~\ref{fig1:means_cartoons}(B) and (C) depicts a smorgasbord of
mathematicized cartoons for simple repression using both thermodynamic and
kinetic models, respectively, that have appeared in previous literature. For
each cartoon, we calculate the fold-change in mean gene expression as predicted
by that model, deferring most algebraic details to
Appendix~\ref{sec:non_bursty}. What we will find is that
for all cartoons the fold-change can be written as a Fermi function 
\begin{equation}
 \text{fold-change} = \left( 1 + \exp(-\Delta F_R + \log(\rho))  \right)^{-1},
 \label{eq:deltaFR_eq_noneq_equiv}
\end{equation}
where the effective free energy contains two terms: the parameters $\Delta F_R$,
an effective free energy parametrizing the repressor-DNA interaction, and
$\rho$, a term derived from the level of coarse-graining used to model all
repressor-free states. In other words, the effective free energy of the Fermi
function can be written as the additive effect of the regulation given by the
repressor via $\Delta F_R$, and the kinetic scheme used to describe the steps
that lead to a transcriptional event via $\log(\rho)$ (See
Figure~~\ref{fig1:means_cartoons}(D), left panel). This implies all models
collapse to a single master curve as shown in
Figure~\ref{fig1:means_cartoons}(D). We will offer some intuition for why this
master curve exists and discuss why at the level of the mean expression, we are
unable to discriminate ``right'' from ``wrong'' cartoons given only measurements
of fold-changes in expression.

@@

Again, we consign all details of the derivation to
Appendix~\ref{sec:non_bursty}. Here we just highlight the general findings for
all five kinetic models. As already shown in Figure~\ref{fig1:means_cartoons}(C)
and (D), all the kinetic models explored can be collapsed onto the master curve.
Given that the repressor-bound state only connects to the rest of the promoter
dynamics via its binding and unbinding rates, $k_R^+$ and $k_R^-$ respectively,
all models can effectively be separated into two categories: a single
repressor-bound state, and all other promoter states with different levels of
coarse graining. This structure then guarantees that, at steady-state, detailed
balance between these two groups is satisfied. What this implies is that the
steady-state distribution of each of the non-repressor states has the same
functional form with or without the repressor, allowing us to write the
fold-change as a product of the ratio of the binding and unbinding rates of the
promoter, and the promoter details. This is \mrm{This is based in one of the
comments from reviewer \# 2. It is empirically true for all models we explored,
but I don't know how to prove that it is a general statement for any
unimaginable complex model.}
\begin{align}
\text{fold-change} &= \left( 1 + \frac{k_R^+}{k_R^-} \rho \right)^{-1},\\
&= (1 + \exp(-\Delta F_R + \log(\rho) ))^{-1},
\end{align}
where $\Delta F_R \equiv -\log(k_R^+/k_R^-)$, and the functional forms of $\rho$
for each model change as shown in Figure~\ref{fig1:means_cartoons}(C). Another
intuitive way to think about these two terms is as follows: in all kinetic
models shown in Figure~1(C) the repressor-bound state can only be reached from a
single repressor-free state. The ratio of these two states --repressor-bound and
adjacent repressor-free state-- must remain the same for all models, regardless
of the details included in other promoter states if $\Delta F_R$ represents an
effective free energy of the repressor binding the DNA operator. The presence of
other states then draws probability density from the promoter being in either of
these two states, making the ratio between the repressor-bound state and
\textit{all} repressor-free states different. The log difference in this ratio
is given by $\log(\rho)$. Since model 1 and model 5 of Figure~1(C) consist of a
single repressor-free state, $\rho$ is then necessarily 1 (See
Appendix~\ref{sec:non_bursty} for further details).

@@

% We have already pointed out that the master curve of~\fig{fig1:means_cartoons}
% is essentially a one-parameter model, the one parameter being $\Delta F_R +
% \log\rho$. By now the reader may be alarmed as to how can we even determine
% $\Delta F_R$ and $\rho$ independently of each other, never mind shedding a lens
% on the internal structure of $\rho$ itself. A hint is provided by the weak
% promoter approximation, invoked repeatedly in prior studies~\cite{Bintu2005c,
% Garcia2011a, Razo-Mejia2018} of simple repression using the 3-state equilibrium
% model in~\fig{fig1:means_cartoons}(B). In that picture, the weak promoter
% approximation means $\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P) \ll 1$,
% meaning therefore $\rho\approx1$.  This approximation can be well justified on
% the basis of the number of RNAP and $\sigma$ factors per cell
% and the strength of binding of RNAP to
% DNA at weak promoters. This is suggestive, but how can we be sure that $\rho$ is
% not, for instance, actually $10^2$ and that $\Delta F_R$ hides a compensatory
% factor? A resolution is offered by an independent inference of $\rho$ in the
% absence of repressors. This was done in~\cite{Razo-Mejia2020} by fitting
% kinetic model 4 in~\fig{fig1:means_cartoons}(C), with zero repressor
% (looking ahead, this is equivalent to model 4
% in Figure~\ref{fig2:constit_cartoons}(A)),
% to single-cell mRNA counts data from~\cite{Brewster2014}. This provided a
% determination of $k^+$ and $k^-$, from which their ratio is estimated to be no
% more than a few $10^{-1}$ and possibly as small as $10^{-2}$.

### 2020-10-14 doc/resubmission/section_05_discussion.tex
implement Rob's changes to main text

Our key insights can be summarized as follows. First, as shown in
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the RNAP interaction with the promoter appear separately and
additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
The fact that thermodynamic and kinetic models for the simple repression motif
can not only be reconciled, but are indistinguishable from each other at the
level of mean mRNA count is a consequence of the possible transitions between
promoter microstates. The promoter microstates in all models shown in
Figure~\ref{figure1:means_cartoons} can explicitly be separated into two groups:
1) promoters with repressor bound, and 2) all other promoters states. The single
path between these two groups of states guarantees detailed-balance between the
groups. We can therefore separate \textit{expression} related transitions, which
by definition must be out of equilibrium, with \textit{regulation} transitions
that can in principle be in equilibrium. This implies that the probability
distribution of the promoter microstates not involving the repressor take the
same functional form, allowing us to write the fold-change as an effective free
energy involving a regulation term, $\Delta F_R$, and a promoter details term,
$\rho$. We offer this argument as a conjecture, and we suspect that a careful
argument using the graph-theoretic framework based on the Matrix-Tree
Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some 
challenges as this theoretical framework applies, at the present, to finite 
graphs.

@@

It is important to mention the potential problems present in all models,
including model 5 from Figure~\ref{fig2:constit_cartoons}(A). Irreversible
reactions as the ones used in all transcription events lead to physical
paradoxes such as infinite entropy production rates. These are mathematical
conveniences rather than physical reality, leaving the inferred rates for the
transcription burst rate and burst size as an effective rate with lots of
(potentially) reversible processes. Also, the state of the art modeling imagines
the promoter existing in a series of discrete states. This assumption might not
be entirely true as effects such as DNA supercoiling~\cite{Chong2014,Sevier2016}
or DNA looping~\cite{Boedicker2013a} can be considered continuous
variables; although there are formal ways of integrating out such degrees of
freedom. Furthermore, when computing the fold-change in gene expression
(Section~\ref{section_02_means}) we assume that the translation efficiency of
mRNA transcript is equal in cells with and without regulation. There might be
some known biological effects that could limit the applicability of this
assumption such as cooperation between the transcription and translation 
machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and 
degradation~\cite{Chen2015} among other potential effects. Another implicit 
assumption in all of the models discussed here is that the effects due to 
changes in gene copy number during the cell cycle have no effects on the 
expression profile. For moderate growth rates as the ones used 
in~\cite{Jones2014} from where we obtained the data, there are at any point at
most two copies of the gene. On a recent publication, we explored the effects
that this change in gene dosage can have in gene expression, finding that the
effects are much more drastic at the protein level~\cite{Razo-Mejia2020}.

>>>>

Our key insights can be summarized as follows. First, as shown in
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the RNAP interaction with the promoter appear separately and
additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
The fact that thermodynamic and kinetic models for the simple repression motif
can not only be reconciled, but are indistinguishable from each other at the
level of mean mRNA count is a consequence of the possible transitions between
promoter microstates. The promoter microstates in all models shown in Figure~1
can explicitly be separated into two groups: 1) promoters with repressor bound,
and 2) all other promoters states. The single path between these two groups of
states guarantees detailed-balance between the groups. We can therefore separate
\textit{expression} related transitions, which by definition must be out of
equilibrium, with \textit{regulation} transitions that can in principle be in
equilibrium. This implies that the probability distribution of the promoter
microstates not involving the repressor take the same functional form, allowing
us to write the fold-change as an effective free energy involving a regulation
term, $\Delta F_R$, and a promoter details term, $\rho$. We offer this argument
as a conjecture, and we suspect that a careful argument using the
graph-theoretic framework based on the Matrix-Tree
Theorem~\cite{Gunawardena2012}, might furnish a ``proof'' not with some
challenges as this theoretical framework applies, at the present, to finite
graphs.

@@

It is important to be clear on the nature of the assumptions  present in all
models, including model 5 from Figure~\ref{fig2:constit_cartoons}(A). First, as
with many of the models used routinely in the analysis of transcription, the
models considered here imagine the promoter as existing in a series of discrete
states. This assumption is clearly an oversimplification due to effects such as
DNA supercoiling~\cite{Chong2014,Sevier2016} or DNA
looping~\cite{Boedicker2013a} which are both parameterized by  continuous
variables describing DNA shape; although there are formal and very interesting
ways of integrating out such degrees of freedom. Another facet of the models
used here is that they can feature irreversible processes characterized by only
a single rate process with no corresponding return pathway at odds with
thermodynamics. These irreversible reactions lead to physical paradoxes such as
infinite entropy production rates. Such models are mathematically convenient and
serve only as an approximation to the more realistic situation in which the
forward rates are much larger than the backward rates. One consequence is that
the inferred rates for the transcription burst rate and burst size  can only be
thought of as effective rate constants. Another idealization of the models
considered here is that when computing the fold-change in gene expression
(Section 2) we assume that the translation efficiency of an mRNA transcript is
the same in cells with and without regulation. There might be some known
biological effects that could limit the applicability of this assumption such as
cooperation between the transcription and translation
machinery~\cite{Proshkin2010}, or coupling of mRNA transcription and
degradation~\cite{Chen2015} among other potential effects. Another implicit
assumption in all of the models discussed here is that the effects due to
changes in gene copy number during the cell cycle have no effects on the
expression profile. For moderate growth rates such as those used
in~\cite{Jones2014} from which we obtained the data, at all points during the
cell cycle there are at most two copies of the gene of interest. In a recent
publication, we explored the effects that this change in gene dosage can have in
gene expression, finding that the effects are much more drastic at the protein
level~\cite{Razo-Mejia2020}.
